---
title: "URD 4: Biased Random Walks"
linestretch: 0.5
output:
 pdf_document:
   latex_engine: xelatex
 html_notebook: default
---

\fontsize{8}{18}

```{r knit_prep, echo=F, results='hide', message=F, warning=F}
library("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE,dev="png",dpi=150)
```

```{r, message=F, warning=F}
library(URD)
```

```{r, include=F}
setwd("~/Dropbox/Jeff-Yiqun/URD-walkthrough/")
```

# Load previous saved object

```{r load-object}
object <- readRDS("obj/object_4_withTips.rds")
```

# Biased Random Walks

### Define parameters of logistic function to bias transition probabilities

```{r}
diffusion.logistic <- pseudotimeDetermineLogistic(object, "pseudotime", optimal.cells.forward=40, max.cells.back=80, pseudotime.direction="<", do.plot=T, print.values=T)
```

### Run walks on cluster

Biased random walks were run on the cluster using the scripts URD-TM.R, URD-TM.sh (to build a biased transition matrix that could be re-used for each tip), and then URD-Walks.R and URD-Walks.sh to parallelize the process of walking from each tip. The commands run by the scripts (if you have a smaller data set that could run on a laptop, for instance) were:

```{r, eval=F, highlight=F}
# Create biased transition matrix
biased.tm <- pseudotimeWeightTransitionMatrix(object, pseudotime = "pseudotime", logistic.params = diffusion.logistic, pseudotime.direction = "<")

# Define the root cells
root.cells <- rownames(object@meta)[object@meta$STAGE=="ZFHIGH"]

# Define the tip cells
tips <- setdiff(unique(object@group.ids[,clustering]), NA)
this.tip <- tips[tip.to.walk] # tip.to.walk was passed by the cluster job array.
tip.cells <- rownames(object@group.ids)[which(object@group.ids[,clustering] == this.tip)]

# Do the random walks
these.walks <- simulateRandomWalk(start.cells=tip.cells, transition.matrix=biased.tm, end.cells=root.cells, n=walks.to.do, end.visits=1, verbose.freq=round(walks.to.do/20), max.steps=5000)

```

### Process walks from cluster

We then load the pre-run walks from the cluster, and process them to determine the visitation frequency of each cell by the walks from each tip. This determines the developmental trajectories, and will be used to determine the branching structure in the data.

```{r}
# Get list of walk files
tip.walk.files <- list.files(path = "walks/dm-8-tm-40-80/", pattern = ".rds", full.names = T)

# Which tips were walked?
tips.walked <- setdiff(unique(object@group.ids$`ZF6S-Cluster-Num`), NA)

# Run through each tip, load the walks, and process them into visitation frequency for that tip.
for (tip in tips.walked) {
  # Get the files for that tip
  tip.files <- grep(paste0("walks-", tip, "-"), tip.walk.files, value=T)
  if (length(tip.files) > 0) {
    # Read the files into a list of lists, and do a non-recursive unlist to combine into one list.
    these.walks <- unlist(lapply(tip.files, readRDS), recursive = F)
    object <<- processRandomWalks(object, walks=these.walks, walks.name=tip, verbose=F)
  }
}
```

# Save objects

```{r}
saveRDS(object, file="obj/object_5_withWalks.rds")
```

